---
title: "Lab5"
author: "Sam Scott"
date: "2026-06-05"
output: html_document
---
Linear Classifiers

#In this module we will build linear classification models specifically applying logistic regression, linear discriminant analysis, penalized logistic regression and nearest shrunken centroids to the Wisconsin breast cancer data set. In the end we will compare the performance of these models on the breast cancer data set.

```{r}
library(caret)
library(pROC)
```


```{r}
#The data set does not have pre set headers so we make them here
cnames <- c('Id', 'Clump_thickness', 'Uniformity_cell_size', 'Uniformity_cell_shape', 
            'Marginal_adhesion', 'Single_e_cell_size', 'Bare_nuclei', 
            'Bland_chromatin', 'Normal_nucleoli', 'Mitoses', 'Class')
#applying the columns and inserting them here
bcancerdata <- read.csv('C:/Users/samsc/Desktop/ADS-503/ADS-503-Predictive-Modeling/breast-cancer-wisconsin.data', header=FALSE, col.names = cnames)
bcancerdata$Bare_nuclei <- as.numeric(as.character(bcancerdata$Bare_nuclei))

```



```{r}
#our response variable is not in factor form which is ideal for discrete variables

bcancerdata$Class <- ifelse(bcancerdata$Class == 4, "malignant", "benign")
bcancerdata$Class <- as.factor(bcancerdata$Class)
table(bcancerdata$Class)

```
```{r}
str(bcancerdata)
```

```{r}
str(bcancerdata)
```

```{r}
summary(bcancerdata)
```

```{r}
corrplot::corrplot(cor(bcancerdata[,1:10]))
set.seed(100)
```

```{r}
trainingRows <- createDataPartition(bcancerdata$Class, p=.8, list=FALSE)
bcancerdata$Class

train <- bcancerdata[trainingRows, ]
test <- bcancerdata[-trainingRows, ]

trainimp <- preProcess(train, "knnImpute")
trainpr <- predict(trainimp, train)
testpr <- predict(trainimp, test)

summary(trainpr)

```

```{r}
summary(testpr)
```

Logistic Regression
#Model log odds as a linear function of the predictor variables

```{r}
ctrl <- trainControl(method = "cv",
                     summaryFunction = twoClassSummary, #ROC metric is associated
                     #with twoClassSummary (area under ROC curve)
                     classProbs = TRUE, #print/show class probabilities
                     savePredictions = TRUE)
set.seed(476)

lrFit <- train(trainpr[,1:10],
               trainpr$Class,
               method = "glm", #generalize linear model
               metric = "ROC",
               trControl = ctrl)
lrFit
```

```{r}
lrFit$finalModel
```

```{r}
lrCM <- confusionMatrix(lrFit, norm = "none")
lrCM
```

```{r}
#plot the ROC curve for the hold-out set
?rev
lrRoc <- roc(response = lrFit$pred$obs,
             predictor = lrFit$pred$malignant,
             levels = rev(levels(lrFit$pred$obs)))
```


```{r}
plot(lrRoc, legacy.axes = TRUE)
```


```{r}
lrRoc$auc
```

```{r}
lrImp <- varImp(lrFit, scale = FALSE)
plot(lrImp)
```

```{r}
testResults <- data.frame(obs = testpr$Class, 
                          LR = predict(lrFit,testpr[,1:10]))
```

Linear Discriminant Analysis
#Goal is to maximize the separability between classes

```{r}
set.seed(476)
ldaFit <- train(trainpr[,1:10],
                trainpr$Class,
                method = "lda",
                preProc = c("center", "scale"),
                metric = "ROC",
                trControl = ctrl)
ldaFit
```

```{r}
ldaFit$finalModel
```

```{r}
ldaCM <- confusionMatrix(ldaFit, norm = "none")
ldaCM
```

```{r}
ldaRoc <- roc(response = ldaFit$pred$obs,
              predictor = ldaFit$pred$malignant,
              levels = rev(levels(ldaFit$pred$obs)))
```

```{r}
plot(ldaRoc, legacy.axes = TRUE)
```

```{r}
varImp(ldaFit)
```

```{r}
testResults$LDA <- predict(ldaFit, testpr[,1:10])
```

Penalized Logistic Regression

```{r}
glmnGrid <- expand.grid(alpha = c(0, .1, .2, .4, .6, .8, 1), lambda = seq(.01, .2, length = 10))
set.seed(476)
glmnFit <- train(trainpr[,1:10],
                trainpr$Class,
                method = "glmnet",
                tuneGrid = glmnGrid,
                preProc = c("center", "scale"),
                metric = "ROC",
                trControl = ctrl)
glmnFit
```

```{r}
glmnFit$results
```

```{r}
plot(glmnFit$finalModel, label = TRUE)
```

```{r}
?coef
coef(glmnFit$finalModel, s = 100) #s corresponds to lambda (lambda value)
```

```{r}
coef(glmnFit$finalModel, s = 0.001) 
```

```{r}
glmnFit$finalModel$lambda
```

```{r}
glmnFit$finalModel$tuneValue

```

```{r}
glmnFit$finalModel$lambdaOpt

```

```{r}
coef(glmnFit$finalModel, s = glmnFit$finalModel$lambdaOpt)

```

```{r}
glmnetCM <- confusionMatrix(glmnFit, norm = "none")
glmnetCM
```
```{r}
## Plot the ROC curve for the hold-out set
glmRoc <- roc(response = glmnFit$pred$obs,
              predictor = glmnFit$pred$malignant,
              levels = rev(levels(glmnFit$pred$obs)))

```

```{r}
plot(glmRoc, legacy.axes = TRUE)
```


```{r}
testResults$glmnet <- predict(glmnFit, testpr[, 1:10])
```

Nearest Shrunken Centroids 
#Typically works well for high dimensional data such as RNA analysis where the number of predictors/features is much larger than number of samples.

```{r}
set.seed(476)
nscFit <- train(trainpr[,1:10],
                trainpr$Class,
                method = "pam",
                preProc = c("center", "scale"),
                tuneGrid = data.frame(threshold = seq(0, 25,length = 30)),
                #threshold is the shrinkage parameter
                metric = "ROC",
                trControl = ctrl) #threshold value is chosen based on ROC

```
```{r}
nscCM <- confusionMatrix(nscFit, norm = "none")
nscCM

```

```{r}
## Plot the ROC curve for the hold-out set
nscRoc <- roc(response = nscFit$pred$obs,
             predictor = nscFit$pred$malignant,
             levels = rev(levels(nscFit$pred$obs)))
```

```{r}
plot(nscRoc, legacy.axes = TRUE)
```

```{r}
testResults$NSC <- predict(nscFit, testpr[,1:10])

```

Model Performance Comparison

```{r}
### Compare Models using ROC curve

plot(lrRoc, type = "s", col = 'red', legacy.axes = TRUE)
plot(ldaRoc, type = "s", add = TRUE, col = 'green', legacy.axes = TRUE)
plot(glmRoc, type = "s", add = TRUE, col = 'blue', legacy.axes = TRUE)
plot(nscRoc, type = "s", add = TRUE, legacy.axes = TRUE)
legend("bottomright", legend=c("LR", "LDA", "GLMNET", "NSC"),
       col=c("red", "green", "blue", "black"), lwd = 2)
title(main = "Compare ROC curves from different models", outer = TRUE)

```

```{r}
confusionMatrix(testResults$LR, testResults$obs, positive = "malignant")
```
```{r}
confusionMatrix(testResults$LDA, testResults$obs, positive = "malignant")

```

```{r}
confusionMatrix(testResults$glmnet, testResults$obs, positive = "malignant")

```

```{r}
confusionMatrix(testResults$NSC, testResults$obs, positive = "malignant")

```