---
title: "M5-A1"
author: "Samuel Scott"
date: "`r Sys.Date()`"
output: html_document
---


```{r include=FALSE}
#optional elapsed time metrics (pairs with last chunk)
library(tictoc)
tic.clearlog()
tic('Total render time')
```

```{r warning=FALSE, message=FALSE}
suppressPackageStartupMessages(library(tidyverse))
library(caret)
library(tidyverse)
library(gt)
library(AppliedPredictiveModeling)
library(pROC)

rseed <- 503
```

Problem 5.1 (60 points)

The hepatic injury data set was described in the introductory chapter and contains 281 unique compounds, each of which has been classified as causing no liver damage, mild damage, or severe damage. These compounds were analyzed with 184 biological screens (i.e., experiments) to assess each compound’s effect on a particular biologically relevant target in the body. The larger the value of each of these predictors, the higher the activity of the compound. In addition to biological screens, 192 chemical fingerprint predictors were determined for these compounds. Each of these predictors represent a substructure (i.e., an atom or combination of atoms within the compound) and are either counts of the number of substructures or an indicator of presence or absence of the particular substructure. The objective of this data set is to build a predictive model for hepatic injury so that other compounds can be screened for the likelihood of causing hepatic injury.

Load the data

```{r 5_1_data}
data(hepatic)
```

Apply the following pre-processing steps:

```{r 5_1_prep}
# Lump all compounds that cause injury into a "Yes" category:
outcome <- tibble(
    any_damage = factor(ifelse(injury == "None", "No", "Yes"),
                        levels = c("Yes", "No") )  #first factor = event of interest
)

#print a table
outcome |> count(any_damage) |> 
    gt() |> 
    grand_summary_rows(columns = n,
                 fns = list(total = ~sum(.))
)
```

The dataframes bio and chem contain the biological assay and chemical fingerprint predictors for the 281 compounds, while the factor variable injury contains the liver damage classification for each compound.

5.1.a (5 points)

Given the size of the dataset and the injury status distribution, describe if you would create a separate training and testing data set

The amount of observations in combination with the class injury status distribution, I would create a training and test data set as I believe there is enough information for a model to learn the training and enough test results to validate training model. The class distribution is not too marginally different enough to see bias in class selection.

5.1.b (5 points)

Which classification statistic would you choose to optimize for this exercise and why?

I would choose the Kappa statistic to optimize as it is great at dealing with class imbalance and specifically used for two class classification. Because our data set has either a 'Yes' or 'No' response, the Kappa statistics would be valuable to optimize due to this data set being two class classification and that the class balance is not exactly even.

Kappa = (O - E) / (1 - E)

5.1.c (20 points)

Perform appropriate pre-processing of data and build logistic regression, linear discriminant analysis, penalized logistic regression and nearest shrunken centroids models described in this chapter for the biological predictors and separately for the chemical fingerprint predictors.

Biological predictors

#Pre process of bio data
```{r 5_1_c_bio}
y <- outcome$any_damage

nzv_bio <- nearZeroVar(bio)
bio <- bio[,-nzv_bio]

correlations <- cor(bio)
highCorr <- findCorrelation(correlations, cutoff = .75)
bio <- bio[,-highCorr]

trainingRows <- createDataPartition(y, p = .75, list=FALSE)

train_bio <- bio[trainingRows,]
test_bio <- bio[-trainingRows,]

y_train <- y[trainingRows]
y_test <- y[-trainingRows]

ctrl <- trainControl(method = "cv",
                     summaryFunction = twoClassSummary,
                     classProbs = TRUE,
                     savePredictions = TRUE)
```

Logistic Regression

```{r}
set.seed(rseed)
log_reg <- suppressWarnings(train(train_bio, y_train,
                 method = "glm",
                 metric = "ROC",
                 trControl = ctrl))
#log_reg
```

Linear Discriminant Analysis 

```{r}
set.seed(rseed)
ldaFit <- train(train_bio,
                y_train,
                method = "lda",
                preProc = c("center", "scale"),
                metric = "ROC",
                trControl = ctrl)
#ldaFit
```

Penalized Logistic Regression

```{r}
glmnGrid <- expand.grid(alpha = c(0, .1, .2, .4, .6, .8, 1), lambda = seq(.01, .2, length = 10))
set.seed(rseed)
glmnFit <- train(train_bio,
                y_train,
                method = "glmnet",
                tuneGrid = glmnGrid,
                preProc = c("center", "scale"),
                metric = "ROC",
                trControl = ctrl)
#glmnFit
```

```{r}
set.seed(rseed)
nscFit <- train(train_bio,
                y_train,
                method = "pam",
                preProc = c("center", "scale"),
                tuneGrid = data.frame(threshold = seq(0, 25,length = 30)),
                #threshold is the shrinkage parameter
                metric = "ROC",
                trControl = ctrl) #threshold value is chosen based on ROC
#nscFit
```

Chemical predictors

```{r 5_1_c_chem}
nzv_chem <- nearZeroVar(chem)
chem <- chem[,-nzv_chem]

correlations2 <- cor(chem)
highCorr2 <- findCorrelation(correlations2, cutoff = .75)
chem <- chem[,-highCorr2]

train_chem <- chem[trainingRows,]
test_chem <- chem[-trainingRows,]
```

Logistic Regression

```{r}
set.seed(rseed)
log_reg2 <- suppressWarnings(train(train_chem, y_train,
                 method = "glm",
                 metric = "ROC",
                 trControl = ctrl))
#log_reg2
```

Linear Discriminant Analysis

```{r}
set.seed(rseed)
ldaFit2 <- suppressWarnings(train(train_chem,
                y_train,
                method = "lda",
                preProc = c("center", "scale"),
                metric = "ROC",
                trControl = ctrl))
#ldaFit2
```

Penalized Logistic Regression

```{r}
set.seed(rseed)
glmnFit2 <- train(train_chem,
                y_train,
                method = "glmnet",
                tuneGrid = glmnGrid,
                preProc = c("center", "scale"),
                metric = "ROC",
                trControl = ctrl)
#glmnFit2
```

Nearest Shrunken Centroid

```{r}
set.seed(rseed)
nscFit2 <- train(train_chem,
                y_train,
                method = "pam",
                preProc = c("center", "scale"),
                tuneGrid = data.frame(threshold = seq(0, 25,length = 30)),
                #threshold is the shrinkage parameter
                metric = "ROC",
                trControl = ctrl) #threshold value is chosen based on ROC
#nscFit2
```


```{r}
train_metrics <- resamples(list(Log_Reg_Bio=log_reg, LDA_Bio=ldaFit, Penalized_Log_Bio =glmnFit,NSC_Bio = nscFit, Log_Reg_Chem=log_reg2, LDA_Chem=ldaFit2, Penalized_Log_Chem=glmnFit2,NSC_Chem = nscFit2))
summary(train_metrics)
```

Which model has the best predictive ability for the biological predictors, and what is the optimal performance? Which model has the best predictive ability for the chemical predictors, and what is the optimal performance? Based on these results, which set of predictors (biological or chemical) contains the most information about hepatic toxicity?

The model with the best predictive ability for biological predictors is the nearest shrunken centroids model with a ROC mean score of 0.55. The model with the best predictive ability for chemical predictors is the penalized logistic regression with a ROC mean score of 0.66. The chemical predictors had a better prediction on determining hepatic toxicity across all models as all models mean ROC scores were higher than the biological models.

5.1.d (5 points)

For the optimal models for both the biological and chemical predictors, what are the top five important predictors?

```{r 5_1_d}
#... code
```

...Answer (optional unless not output by table)

5.1.e (20 points)

Combined predictors

Now combine the biological and chemical fingerprint predictors into one predictor set. Retrain the same set of predictive models you built from part (c).

```{r 5_1_e}
#... code
```

Which model yields the best predictive performance?

...Answer...

Is the model performance better than either of the best models from part (c)?

...Answer...

What are the top five important predictors for the optimal model?

```{r 5_1_e_imp}
#...code
```

How do these compare with the optimal predictors from each individual predictor set?

...Answer...

5.1.f (5 points)

Which model (either model of individual biology or chemical fingerprints or the combined predictor model), if any, would you recommend using to predict compounds’ hepatic toxicity? Explain.

...Answer...

Problem 5.2 (30 points)

Brodnjak-Vonina et al. (2005) develop a methodology for food laboratories to determine the type of oil from a sample. In their procedure, they used a gas chromatograph (an instrument that separates chemicals in a sample) to measure seven different fatty acids in an oil. These measurements would then be used to predict the type of oil in a food sample. To create their model, they used 96 samples of seven types of oils. These data can be found in the caret package using data(oil). The oil types are contained in a factor variable called oilType. The types are pumpkin (coded as A), sunflower (B), peanut (C), olive (D), soybean (E), rapeseed (F), and corn (G). We would like to use these data to build a model that predicts the type of oil-based on a sample’s fatty acid percentages.

```{r 5_2_data}
data(oil)
table(oilType)
```

5.2.a (5 points)

Like the hepatic injury data, these data suffer from imbalance. Given this imbalance, should the data be split into training and test sets?

...Answer...

5.2.b (5 points)

Which classification statistic would you choose to optimize for this exercise and why?

...Answer...

5.2.c (20 points)

Build linear discriminant analysis, penalized multinomial regression, and nearest shrunken centroids models to this data; predict the fitted models on the training data set and evaluate (show the confusion matrix for each model output) ...

(Warnings suppressed - these warnings related to the sparsity of the data for some oil types.)

LDA

```{r 5_2_c-lda}
#...code
```

PLR

```{r 5_2_c-plr}
#...code
```

NSC

```{r 5_2_c-nsc}
#...code
```

Accuracies Summary

```{r}
# summarize the accuracies for all models in a table
```

which model performs best on these data?

...Answer...

Which oil type does the optimal model most accurately predict?

...Answer...

Which oil type does the optimal model least accurately predict?

...Answer...

Elapsed times

{r include=FALSE}
toc(log = TRUE, quiet = TRUE)
tic_times <- tic.log() |> unlist() |> as_tibble()
tic_times |> gt()

