---
title: "M6-A1"
author: "Sam Scott"
date: "2026-06-14"
output: html_document
---
```{r}
library(caret)
library(kknn)
library(lsa) #cosine similarity
library(e1071) #naive bayes
library(caTools)
seed <- 756
```


#1.KNN dataset

```{r}
knn_train <- data.frame(
  x1 = c(1:6),
  x2 = c(2:7),
  x3 = c(30, 40, 50, 60, 70, 80),
  Class = c('A', 'B', 'A', 'B','B', 'A')
)

knn_test <- data.frame(
  x1 = c(9,7,8),
  x2 = c(4,8,7),
  x3 = c(15,45,75)
)
```

#a) Perform the analysis WITHOUT doing any preprocessing of the data. (5 points)

```{r}
#no resampling
no_resample <- trainControl(method = "none")
set.seed(seed)
knn <- train(Class ~ .,
                  knn_train,
                  method = "knn",
                  tuneGrid = data.frame(k = 3),
                  trControl = no_resample
                  )
pred_wthout <- predict(knn, knn_test)
pred_wthout
```
#b)	Perform the analysis WITH appropriate preprocessing of the data. (8 points)

```{r}
#no resampling
no_resample <- trainControl(method = "none")
set.seed(seed)
knn2 <- train(Class ~ .,
                  knn_train,
                  method = "knn",
                  tuneGrid = data.frame(k = 3),
                  preProcess = c("center", "scale"),
                  trControl = no_resample
                  )
pred_wth <- predict(knn2, knn_test)
pred_wth
```
#c) Do you observe any differences between the predictions with and without preprocessing? (2 points)

The differences between the predictions with and without preprocessing occur in the first test sample (9, 4, 15). The without preprocess knn model predicted class A while the preprocess knn model predicted class B.  


#2. Email spam filtering problem

```{r}
train_bag <- data.frame(
            money = c(3,1,0,0,0),
            free = c(0,2,0,0,1),
            `for` = c(0,1,1,1,0),
            gambling = c(0,1,1,0,0),
            fun = c(0,1,1,3,0),
            machine = c(0,0,0,1,1),
            learning = c(0,0,0,1,1),
            spam = c(TRUE, TRUE, TRUE, FALSE,  FALSE)
            )
#train_bag
```
#a)	What target level would a nearest neighbor model using Euclidean distance return for the following email: “machine learning for free”? (10 points)

```{r}
test_bag <- data.frame(
  money = 0,
  free = 1,
  `for` = 1,
  gambling = 0,
  fun = 0,
  machine = 1,
  learning = 1
)
#test
```

```{r}
#made spam a factor
train_bag$spam<-as.factor(train_bag$spam)
#train_bag$spam
set.seed(seed)
bagofwordsmodel <- kknn(spam ~ ., train_bag, test_bag, k=1, distance = 2)

#Look up fitted.values for predicting on kknn
pred_bag <- bagofwordsmodel$fitted.value
pred_bag
```
#b)	What target level would a k-NN model with k = 3 and using Euclidean distance return for the same query? (5 points)

```{r}
set.seed(seed)
bagofwordsmodel3 <- kknn(spam ~ ., train_bag, test_bag, k=3, distance = 2)

pred_bagb <- bagofwordsmodel3$fitted.value
pred_bagb
```
The bagofwordsmodel with k=3 would also return False as a Spam email.

#c)	What target level would a k-NN model with k = 3 and using Manhattan distance return for the same query? (5 points)

```{r}
set.seed(seed)
bagofwordsmodel_manhat <- kknn(spam ~ ., train_bag, test_bag, k=3, distance = 1)

pred_bagc <- bagofwordsmodel_manhat$fitted.value
pred_bagc
```
#d)	There are a lot of zero entries in the spam bag-of-words dataset. This is indicative of sparse data and is typical for text analytics. Cosine similarity is often a good choice when dealing with sparse non-binary data. What target level would a 3-NN model using cosine similarity return for the query? (10 points)


```{r}
#https://www.geeksforgeeks.org/r-language/how-to-calculate-cosine-similarity-in-r/

money =  c(3,1,0,0,0)
free = c(0,2,0,0,1)
`for` = c(0,1,1,1,0)
gambling = c(0,1,1,0,0)
fun = c(0,1,1,3,0)
machine = c(0,0,0,1,1)
learning = c(0,0,0,1,1)
#test
test_bag_vector <- c(0,1,1,0,0,1,1)

cos_sim <- cbind(money, free, `for`, gambling, fun, machine, learning, test_bag_vector)
cos_calculations <- cosine(cos_sim)
training_emails <- cos_calculations[,ncol(cos_calculations)][1:5]
#str(training_emails)
three_closest_neighbors <-order(training_emails, decreasing = TRUE)[1:3]
bool_vector  <- train_bag$spam[three_closest_neighbors]
#bool_vector
numeric_vector <-as.numeric(bool_vector)
#numeric_vector
prediction_cos_sim <- ifelse(sum(numeric_vector) >= 2, TRUE, FALSE)
prediction_cos_sim
```


#3) Predictive data analytics models are often used as tools for process quality control and fault detection. The task in this question is to create a naive Bayes model to monitor a waste water treatment plant. The table below lists a dataset containing details of activities at a waste water treatment plant for 14 days. Each day is described in terms of six descriptive features that are generated from different sensors at the plant. SS-IN measures the solids coming into the plant per day; SED-IN measures the sediment coming into the plant per day. Features SS-OOUT, SED-OUT, and COND-OUT are the corresponding measurements for the water flowing out of the plant. The target feature, STATUS, reports the current situation at the plant:

#ok, everything is working correctly
#settler, there is a problem with the plant settler equipment
#solids, there is a problem with the amount of solids going through the plant

```{r}
waste_plant <- data.frame(
  ID = c(1:13),
  `SS-IN` = c(168, 156, 176, 256, 230, 116, 242,242,174, 1004, 1228, 964, 2008),
  `SED-IN` = c(3,3,3.5,3,5,3,7,4.5,2.5,35,46,17,32),
  `COND-IN` = c(1814,1358,2200,2070,1410, 1238,1315,1183,1110,1218,1889,2120,1257),
  `SS-OUT` = c(15,14,16,27,131,104,104,78,73,81,82.4,20,13),
  `SED-OUT` = c(0.001, 0.01,0.005,0.2,3.5,0.06,0.01,0.02,1.5,1172,1932,1030,1038),
  `COND-OUT` = c(1879,1425,2140,2700,1575,1221,1434,1374,11256,33.3,43.1,1966,1289),
  `STATUS` = c("ok", "ok", "ok", "ok", "settler", "settler", "settler", "settler", "settler", "solids", "solids", "solids", "solids")
)
waste_plant
```

#a)	Create a naive Bayes model that uses probability density functions to model the descriptive features in this dataset (assume that all the descriptive features are normally distributed). Show clearly the distribution of each of the predictors conditional on status and the prior probabilities of the status. (15 points)

```{r}
#https://www.geeksforgeeks.org/r-language/naive-bayes-classifier-in-r-programming/
set.seed(seed)
dim(waste_plant)

#id gives no predictive power
waste_plant <- waste_plant[,-1]

#FACTOR CLASS
waste_plant$STATUS <- as.factor(waste_plant$STATUS)
str(waste_plant$STATUS)

#GET PDFS using tapply:

#SS.IN
ssin_mean <- tapply(waste_plant$SS.IN, waste_plant$STATUS, mean)
ssin_sd <-tapply(waste_plant$SS.IN, waste_plant$STATUS, sd)
#SED.IN
sedin_mean <- tapply(waste_plant$SED.IN, waste_plant$STATUS, mean)
sedin_sd <- tapply(waste_plant$SED.IN, waste_plant$STATUS, sd)
#COND.IN
condin_mean <- tapply(waste_plant$COND.IN, waste_plant$STATUS, mean)
condin_sd <- tapply(waste_plant$COND.IN, waste_plant$STATUS, sd)
#SS.OUT
ssout_mean <- tapply(waste_plant$SS.OUT, waste_plant$STATUS, mean)
ssout_sd <- tapply(waste_plant$SS.OUT, waste_plant$STATUS, sd)
#SED.OUT
sedout_mean <- tapply(waste_plant$SED.OUT, waste_plant$STATUS, mean)
sedout_sd <- tapply(waste_plant$SED.OUT, waste_plant$STATUS, sd)
#COND.OUT
condout_mean <- tapply(waste_plant$COND.OUT, waste_plant$STATUS, mean)
condout_sd <- tapply(waste_plant$COND.OUT, waste_plant$STATUS, sd)

prob_dens_func <- cbind(
  `SS-IN-mean` = ssin_mean,
  `SS-IN-sd` = ssin_sd,
  `SED-IN-mean` = sedin_mean,
  `SED-IN-sd` = sedin_sd,
  `COND-IN-mean` = condin_mean,
  `COND-IN-sd` = condin_sd,
  `SS-OUT-mean` = ssout_mean,
  `SS-OUT-sd` = ssout_sd,
  `SED-OUT-mean` = sedout_mean,
  `SED-OUT-sd` = sedout_sd,
  `COND-OUT-mean` = condout_mean,
  `COND-OUT-sd` = condout_sd
)
prob_dens_func
#nrow(waste_plant)
prior_probs <- table(waste_plant$STATUS) / nrow(waste_plant) * 100
prior_probs

#Model also shows prior probs and pdfs
nb_wp <- naiveBayes(STATUS ~ ., waste_plant)
#nb_wp
```
#b)	What prediction will the naive Bayes model return for the following query? SS-IN = 222, SED-IN = 4.5, COND-IN = 1518, SS-OUT = 74, SED-OUT = 0.25, COND-OUT = 1642. (10 points)

```{r}
nb_func <- function(ssin, sedin, condin, ssout, sedout, condout) {
  naive_vals<- data.frame(
    `SS-IN` = ssin,
    `SED-IN` = sedin,
    `COND-IN` = condin,
    `SS-OUT` = ssout,
    `SED-OUT` = sedout,
    `COND-OUT` = condout
  )
  predict(nb_wp, naive_vals)
}
nb_func(222, 4.5, 1518, 74, 0.25, 1642)
```
#Settler is what the naive-bayes model predicted on the values given

