---
title: "M6-A1"
author: "Sam Scott"
date: "2026-06-14"
output: html_document
---
```{r}
library(caret)
```


#1.KNN dataset

```{r}
knn_train <- data.frame(
  x1 = c(1:6),
  x2 = c(2:7),
  x3 = c(30, 40, 50, 60, 70, 80),
  Class = c('A', 'B', 'A', 'B','B', 'A')
)

knn_test <- data.frame(
  x1 = c(9,7,8),
  x2 = c(4,8,7),
  x3 = c(15,45,75)
)
```

#a) Perform the analysis WITHOUT doing any preprocessing of the data. (5 points)

```{r}
#no resampling
no_resample <- trainControl(method = "none")

knn <- train(Class ~ .,
                  knn_train,
                  method = "knn",
                  tuneGrid = data.frame(k = 3),
                  trControl = no_resample
                  )
pred_wthout <- predict(knn, knn_test)
pred_wthout
```
#b)	Perform the analysis WITH appropriate preprocessing of the data. (8 points)

```{r}
#no resampling
no_resample <- trainControl(method = "none")

knn2 <- train(Class ~ .,
                  knn_train,
                  method = "knn",
                  tuneGrid = data.frame(k = 3),
                  preProcess = c("center", "scale"),
                  trControl = no_resample
                  )
pred_wth <- predict(knn2, knn_test)
pred_wth
```
#c) Do you observe any differences between the predictions with and without preprocessing? (2 points)

The differences between the predictions with and without preprocessing occur in the first test sample (9, 4, 15). The without preprocess knn model predicted class A while the preprocess knn model predicted class B.  



