---
title: "M6-A1"
author: "Sam Scott"
date: "2026-06-14"
output: html_document
---
```{r}
library(caret)
library(kknn)
library(lsa)
```


#1.KNN dataset

```{r}
knn_train <- data.frame(
  x1 = c(1:6),
  x2 = c(2:7),
  x3 = c(30, 40, 50, 60, 70, 80),
  Class = c('A', 'B', 'A', 'B','B', 'A')
)

knn_test <- data.frame(
  x1 = c(9,7,8),
  x2 = c(4,8,7),
  x3 = c(15,45,75)
)
```

#a) Perform the analysis WITHOUT doing any preprocessing of the data. (5 points)

```{r}
#no resampling
no_resample <- trainControl(method = "none")

knn <- train(Class ~ .,
                  knn_train,
                  method = "knn",
                  tuneGrid = data.frame(k = 3),
                  trControl = no_resample
                  )
pred_wthout <- predict(knn, knn_test)
pred_wthout
```
#b)	Perform the analysis WITH appropriate preprocessing of the data. (8 points)

```{r}
#no resampling
no_resample <- trainControl(method = "none")

knn2 <- train(Class ~ .,
                  knn_train,
                  method = "knn",
                  tuneGrid = data.frame(k = 3),
                  preProcess = c("center", "scale"),
                  trControl = no_resample
                  )
pred_wth <- predict(knn2, knn_test)
pred_wth
```
#c) Do you observe any differences between the predictions with and without preprocessing? (2 points)

The differences between the predictions with and without preprocessing occur in the first test sample (9, 4, 15). The without preprocess knn model predicted class A while the preprocess knn model predicted class B.  


#2. Email spam filtering problem

```{r}
train_bag <- data.frame(
            money = c(3,1,0,0,0),
            free = c(0,2,0,0,1),
            `for` = c(0,1,1,1,0),
            gambling = c(0,1,1,0,0),
            fun = c(0,1,1,3,0),
            machine = c(0,0,0,1,1),
            learning = c(0,0,0,1,1),
            spam = c(TRUE, TRUE, TRUE, FALSE,  FALSE)
            )
#train_bag
```
#a)	What target level would a nearest neighbor model using Euclidean distance return for the following email: “machine learning for free”? (10 points)

```{r}
test_bag <- data.frame(
  money = 0,
  free = 1,
  `for` = 1,
  gambling = 0,
  fun = 0,
  machine = 1,
  learning = 1
)
#test
```

```{r}
#made spam a factor
train_bag$spam<-as.factor(train_bag$spam)
#train_bag$spam

bagofwordsmodel <- kknn(spam ~ ., train_bag, test_bag, k=1, distance = 2)

#Look up fitted.values for predicting on kknn
pred_bag <- bagofwordsmodel$fitted.value
pred_bag
```
#b)	What target level would a k-NN model with k = 3 and using Euclidean distance return for the same query? (5 points)

```{r}
bagofwordsmodel3 <- kknn(spam ~ ., train_bag, test_bag, k=3, distance = 2)

pred_bagb <- bagofwordsmodel3$fitted.value
pred_bagb
```
The bagofwordsmodel with k=3 would also return False as a Spam email.

#c)	What target level would a k-NN model with k = 3 and using Manhattan distance return for the same query? (5 points)

```{r}
bagofwordsmodel_manhat <- kknn(spam ~ ., train_bag, test_bag, k=3, distance = 1)

pred_bagc <- bagofwordsmodel_manhat$fitted.value
pred_bagc
```
#d)	There are a lot of zero entries in the spam bag-of-words dataset. This is indicative of sparse data and is typical for text analytics. Cosine similarity is often a good choice when dealing with sparse non-binary data. What target level would a 3-NN model using cosine similarity return for the query? (10 points)


```{r}
#https://www.geeksforgeeks.org/r-language/how-to-calculate-cosine-similarity-in-r/

money =  c(3,1,0,0,0)
free = c(0,2,0,0,1)
`for` = c(0,1,1,1,0)
gambling = c(0,1,1,0,0)
fun = c(0,1,1,3,0)
machine = c(0,0,0,1,1)
learning = c(0,0,0,1,1)
#test
test_bag_vector <- c(0,1,1,0,0,1,1)

cos_sim <- cbind(money, free, `for`, gambling, fun, machine, learning, test_bag_vector)
cos_calculations <- cosine(cos_sim)
training_emails <- cos_calculations[,ncol(cos_calculations)][1:5]
#str(training_emails)
three_closest_neighbors <-order(training_emails, decreasing = TRUE)[1:3]
bool_vector  <- train_bag$spam[three_closest_neighbors]
#bool_vector
numeric_vector <-as.numeric(bool_vector)
#numeric_vector
prediction_cos_sim <- ifelse(sum(numeric_vector) >= 2, TRUE, FALSE)
prediction_cos_sim
```
