---
title: "R_Data_Types"
author: "Sam Scott"
date: "2026-05-09"
output: html_document
---

#Set our work directory to be in our ADS-503 module
setwd("C:/Users/samsc/Desktop/ADS-503/ADS-503-Predictive-Modeling")


```{r}

xn <- c(1,2,3) # numeric vector
yn = c(1,2,3)

str(xn) # structure of the R object
print(str(xn)) # apparently the structure of the R object is NULL
class(xn) #numeric class

xc <- c('a', 'b', 'c') # character vector
str(xc)
class(xc)

xm <- matrix(1:20, nrow=5, ncol=4) #matrix of size 20 having 5 rows and 4 columns like so
# 0 1 2 3 4
# 5 6 7 8 9
# 10 11 12 13 14
# 15 16 17 18 19

# I was wrong about the structure it is 
 #[,1] [,2] [,3] [,4]
#[1,]    1    6   11   16
#[2,]    2    7   12   17
#[3,]    3    8   13   18
#[4,]    4    9   14   19
#[5,]    5   10   15   20

xm # so matrices in R populate the row first then the columns 
# also R starts with 1 and Python starts with 0 so R matrix in this 
# example 

?matrix #question mark helps you get information on a topic
# in this case we want to get information on the matrix topic

populate_cols_first <- matrix(1:20, nrow=5, ncol=4, byrow=TRUE) #byrow default is false. This parameter tells you how the matrix will be populated. If True populate cols first otherwise populate rows first
populate_cols_first 

str(populate_cols_first)
class(populate_cols_first)

?list
?exp
x1 <- list(a = 1:10, beta = exp(-3:3), logic = c(TRUE, FALSE,FALSE, TRUE)) #my guess is x1 has three different lists. the first one (a) is 
# a standard list populating 1 to 10 with 1 row and 10 columns. The 
# beta list is a list of the exponents of 3s (not sure) and the logic list is a boolean mask list of itself?

x1$a #$ is used to return single columns


x1[2] #gets the beta list which is a list of exp values from -3 to 3
# so exp(-x) = 1/exp(x) so exp(3) does not equal exp(-3)

x1[[2]]

str(x1)
class(x1)
?letters #built in constant to putting letters in a list
?rep
#Dataframes
xd1 <- data.frame(x = 1:10, y = letters[1:10], z= rep(c(TRUE, FALSE), 5), stringsAsFactors = FALSE) 
# x create a dataframe 1 to 10 single row 10 columns
# y is the same as x
# z not sure what rep does
# stringsASFactors make sure that no string types are factor types or
# discrete variables

str(xd1)

# x creates a dataframe 1 to 10 single row 10 columns
# y creates a letter dataframe first 10 letters in lowercase single row 10 columns
# z creates the (TRUE FALSE) vector we made 5 times so 10 total values
# single row 10 columns

head(xd1)

##   x y     z
## 1 1 a  TRUE
## 2 2 b FALSE
## 3 3 c  TRUE
## 4 4 d FALSE
## 5 5 e  TRUE
## 6 6 f FALSE

names(xd1) # names of columns in data frame
#[1] "x" "y" "z"                                           

dim(xd1) # dimension of data frame

# 10 rows by 3 columns
                                                          
nrow(xd1) # number of rows of data frame
#10

ncol(xd1) # number of cols of data frame
#3

xd1[1, ] #first row

xd1[, 1] # first column

xd1[2, 3] # second row third column

xd1[, c("y", "z")] # dataframe columns y and z

xd1[xd1$x > 5, ] # subset data frame
# this will essentially give you a dataframe with value greater than 5
# in the x column
#:
# 6, f, FALSE
# 7, g, TRUE
# 8, h, FALSE
# 9, i, TRUE
# 10, j, FALSE
# DONT FORGET IMPLIED INDEX WITH DATAFRAMES!!!!!

xds <- subset(xd1, x > 5) # subset data frame using subset function
xds # same as line 116 

subset(xd1, x > 5 & z == TRUE) #get values greater than 5 in the x #column and also get the values that are set to TRUE in the z column

xds1 <- subset(xd1, x > 5, select = c(x,y)) # subset and select data frame
# my assumption is that this will return x and y columns that are greater than 5 in x column

xds2 <- subset(xd1, x>5, select = -z) 
# returns everything that is not the z column greater than 5 in x column. In other words xds2 = xds1

```
#Reading and writing files in R
```{r}

getwd() #get working directory
setwd("C:/Users/samsc/Desktop/ADS-503/ADS-503-Predictive-Modeling")
cereals <- read.csv("C:/Users/samsc/Desktop/ADS-503/ADS-503-Predictive-Modeling/cereals.CSV")
str(cereals)

cereals2 <- read.csv("C:/Users/samsc/Desktop/ADS-503/ADS-503-Predictive-Modeling/cereals.CSV", stringsAsFactors = FALSE)
str(cereals2)

write.csv(xds2, file= 'xds2.csv') #Creating a csv


```
#Apply functions in R
```{r}
data(cars)
head(cars)
cars

dim(cars) # dimensions of cars data set

apply(cars, 1, mean) # row mean
#row 1: (4 + 2) / 2
#row 2: (4 + 10)/ 2
#row 3: (7 + 4)/ 2
#etcetera...

apply(cars, 2, mean) # column mean

?airquality # Daily air quality measurements in New York, May to September 1973

data(airquality)
head(airquality)

?lapply 
aq_class <- lapply(airquality, class) # get class of each column in the dataframe
aq_class

#numeric is float and integer is integer

aq_mean <- lapply(airquality, mean) #get mean of each column in the dataframe
aq_mean

aq_mean_cleaned <- lapply(airquality, mean, na.rm= TRUE) # get mean of each column in the data frame with NA's removed
aq_mean_cleaned

airquality_s <- airquality[, lapply(airquality, class) == 'integer'] #subset data frame using lapply function
#this should remove the wind section which is of type numeric while the rest of the categories are integers

?sapply #simplified version of lapply?

sapply(airquality, function(x) sum(is.na(x))) #using sapply to find NA's

aq_mean_cleaned_vector <- sapply(airquality, mean, na.rm = TRUE)
aq_mean_cleaned_vector

names(aq_mean_cleaned_vector) #column names of data set

```
#Some useful functions in R for module 1

``` {r}
data(iris)
str(iris)


pairs(iris) # pairwise scatter plots

cor(iris)  # correlation between predictors, this will give error because the fifth columns our target
#variable is classification variable which is nonnumeric it is string or factor

cor(iris[,-5]) # correlation, removes the target variable from correlation

boxplot(iris)

?par #query graph parameters
??mfrow

par(mfrow = c(1,2)) #draw two plots side by side 1 row 2 columns
boxplot(iris$Sepal.Length ~ iris$Species)
boxplot(iris$Sepal.Width ~ iris$Species)

par(mfrow = c(1,3)) 
hist(iris$Sepal.Length)
hist(iris$Sepal.Width)
hist(iris$Petal.Length)

install.packages("Hmisc")
library(Hmisc)


hist.data.frame(iris) #get the whole graphical info of each attribute

hist.data.frame(airquality) #very cool function


```
#Box Cox Transformations
#Boxcox transformations are used to transform a predictor variable with non-normal distribution into predictor with normal distribution.
```{r}
set.seed(20)
?rgamma #non-normal distribution graph
x <- rgamma(1000, 2)
x
hist(x)

library(e1071)
skewness(x) #1.155639 recall skewness?

y <- rnorm(1000)
hist(y)
skewness(y) #0.05625155

library(caret)

bct <- BoxCoxTrans(x)
x_trans <- predict(bct, x)
hist(x_trans)
skewness(x_trans) #-0.1014337

z <- c(x, 0)
bct2 <- BoxCoxTrans(z)
z <- z + 10^-4
bct2 <- BoxCoxTrans(z)
bct
```
#PCA 
#It is commonly used for dimensionality reduction by projecting each data point onto only the first few principal components to obtain lower-dimensional data while preserving as much of the data’s variation as possible.

#Principal components are directions that have maximum variance when data is projected onto them and at the same time are uncorrelated with each other.

#Principal components are linear combinations of original predictors.

# Example 1

set.seed(100)
?rnorm
x <- rnorm(1000)
y <- rnorm(1000)
z <- rnorm(1000)

m <- matrix(data= c(x,y,z), nrow = 1000, ncol = 3)

m_pca <- prcomp(m) #pca method prcomp
str(m_pca)

m_pca_var <- m_pca$sdev^2 #we get pca variance to #because that is PCA's goal when reducing dimension, #get the most variance of the data
m_pca_var * 100/sum(m_pca_var) #checking to see how #much of that variance pca got

??rotation
 
m_pca_var
m_pca$rotation #how strongly each variable is contributing via loadings or weights

m_pca$x[1:20,] 

#Example 2
t <- (x + y + z)/3
m1 <- matrix(data = c(x,y,z,t), nrow = 1000, ncol = 4)

m1_pca <- prcomp(m1)
m1_pca_var <-m1_pca$sdev^2
m1_pca_var * 100/sum(m1_pca_var)

?biplot #a type of exploratory graph used in multivariate statistics to visualize both the observations (samples) and #variables (features) of a data matrix simultaneously
biplot(m1_pca, scale=0) #plots numbers between 1 and 1000
biplot(m1_pca, choices = 2:3, scale = 0)


